The purpose of this script is to prepare willmore data

Main outputs we want
1. List of independent detections to 30 minutes threshold
2. Camera deployment file
3. Site covariate file

Andrew Barnas
April 9th 2024
andrew.f.barnas@gmail.com

Load in packages we will need, setup working directories, and load in files
```{r}
#Clear everything out and start fresh
rm(list=ls())

library(tidyr)        # data tidying
library(stringr)      # working with strings
library(dplyr)        # data manipulation
library(reproducible) # reproducible working directories
library(ggplot2)      # plots
library(ggpubr)       # combining plots
library(lubridate)   # dates


#Setup input and output paths
input_directory <- checkPath(file.path(getwd(), "inputs"), 
                            create = TRUE)

output_directory <- checkPath(file.path(getwd(), "outputs"), 
                             create = TRUE)


#Read in the camera detections. Now Willmore is a bit strange because there is a seperate grid for the winter and summer regions, so we will have to read these in seperately and figure out how we would like to merge them together. Probably some different datetime and other formats as well.
dets_summer<-read.csv(file.path(input_directory, "WillmoreSummer_alldata.csv"), header = TRUE)
dets_winter<-read.csv(file.path(input_directory, "WillmoreWinter_alldata.csv"), header = TRUE)

#Need something about camera deployment in here too
cams<-read.csv(file.path(input_directory, "SUMMER WILLMORE BIODIVERSITY SITE LOCATIONS CORRECTED 2017.csv"), header = TRUE)


```


```{r}
#I am intentionally dealing with the summer and winter data seperately first, as these operated on different years, so they may have different errors and formats to deal with. I will combine them at the end!

#First I am curious about how many different sites there are
n_distinct(dets_summer$Location)
n_distinct(dets_winter$Location)

#Ok so according to above, I have detections from 119 cameras, but location information from 124 cameras
#This could be due to camera errors and whatnot, will deal with this a bit later
n_distinct(cams$Camera_Site)

#Looks like there is a blank, lets remove Locations without names
unique(dets_summer$Location)
dets_summer<-dets_summer%>%
  filter(Location != "")

#Same thing, there is a blank
unique(dets_winter$Location)
dets_winter<-dets_winter%>%
  filter(Location != "")


#Do any of these camera names overlap at all? Nope
unique(dets_summer$Location) %in% unique(dets_winter$Location)

#Lets clean up the detection file so that we only keep things we want
colnames(dets_summer)
colnames(dets_winter)

dets_summer<-dets_summer%>%
  dplyr::select(Date, Time, Location, Species)
dets_winter<-dets_winter%>%
  dplyr::select(Date, Time, Location, Species)

```


Ok lets create the camera operability matrix the array
```{r}
#Ok lets create the camera operability matrix. To do this we need to first know the first and last dates of operation for each camera. In the Willmore data we do not have a timelapse funciton, so this depends on knowing when the first and last detection (trigger) was.

#First read in the datetimes for the detections
dets_summer<-dets_summer%>%
  mutate(datetime = paste(Date, Time, sep = " "),
         datetime = dmy_hm(datetime))

#quick look at the camera operability for these cameras
dets_summer%>%
  group_by(Location)%>%
  mutate(start_date = min(datetime),
         end_date = max(datetime))%>%
  slice(1)%>%
  ggplot(aes(y = Location))+
  geom_segment(aes(x = start_date, xend = end_date, y = Location, yend = Location))

#Camera SW-E5 looks a bit strange...very long deployment. Going to check it out in more detail
A<-dets_summer%>%
  filter(Location == "SW-E5")

#Yeah so looks to me like there is an errant observation in July 2010 that is throwing things off?
ggplot(A, aes(x = datetime))+
  geom_histogram()

#I am going to make an executive decision here that this must be an error. What are the chances of this camera having zero obsverations from late 2009 until a random wolverine on July 23rd 2010?
A%>%
  filter(datetime > ymd("2009-12-31"))
#Removing it here
dets_summer<-dets_summer%>%
  filter(!(Location == "SW-E5" & Date == "23-Jul-10"))

#This looks better.
dets_summer%>%
  group_by(Location)%>%
  mutate(start_date = min(datetime),
         end_date = max(datetime))%>%
  slice(1)%>%
  ggplot(aes(y = Location))+
  geom_segment(aes(x = start_date, xend = end_date, y = Location, yend = Location))

#This camera appears to only have one day of operability...likely a camera error, going to remove it.
dets_summer%>%
  filter(Location == "SW-S27")

dets_summer<-dets_summer%>%
  filter(Location != "SW-S27")

```


############################################
```{r}

#Now the winter cameras, one quick thing to note, Site W55? needs to be renamed to W35 accoring to the readme
unique(dets_winter$Location)
dets_winter<-dets_winter%>%
  mutate(Location = gsub(pattern = "WW-W55?", replacement = "WW-W35", Location, fixed = TRUE))
unique(dets_winter$Location)

#Now the winter cameras
dets_winter<-dets_winter%>%
  mutate(datetime = paste(Date, Time, sep = " "),
         datetime = dmy_hm(datetime))

#Lets take a look at this operability matrix
dets_winter%>%
  group_by(Location)%>%
  mutate(start_date = min(datetime),
         end_date = max(datetime))%>%
  slice(1)%>%
  ggplot(aes(y = Location))+
  geom_segment(aes(x = start_date, xend = end_date, y = Location, yend = Location))

#Everything looks mostly good except for WW-W2, lets take a closer look
A<-dets_winter%>%
  filter(Location == "WW-W2")
#Yeah strange, looks to be some outliers into the summer and fall??

ggplot(A, aes(x = datetime))+
  geom_histogram()
#Yeah check these out, a bit strange. Going to remove them.
A%>%
  filter(datetime > ymd("2007-05-01"))
dets_winter<-dets_winter%>%
  filter(!(Location == "WW-W2" & dmy(Date) > ymd("2007-05-01")))

#Looks better
dets_winter%>%
  group_by(Location)%>%
  mutate(start_date = min(datetime),
         end_date = max(datetime))%>%
  slice(1)%>%
  ggplot(aes(y = Location))+
  geom_segment(aes(x = start_date, xend = end_date, y = Location, yend = Location))

#Ok so now I think we can take a first crack at a camera operability matrix.
dets<-rbind(dets_summer, dets_winter)

dets<-dets%>%
  group_by(Location)%>%
  mutate(start_date = min(datetime),
         end_date = max(datetime))%>%
  slice(1)%>%
  dplyr::select(Location, start_date, end_date)%>%
  mutate(n_days = difftime(end_date, start_date, units = "days"))

#Error checking here, want to see distribution of days
#Looks fine. 
ggplot(dets, aes(x = n_days))+
  geom_histogram()

#Ok last thing, lets add the coordinates into the detection file.
#First the naming conventions are off, need to fix. 
#Going to add SW or WW to the camera deployment files
unique(dets$Location)
unique(cams$Camera_Site)


dets<-dets%>%
 rename(site = Location)

cams<-cams%>%
  mutate(site = case_when(Season == "Winter" ~ paste("WW-", Camera_Site, sep = ""),
                          Season == "Summer" ~ paste("SW-", Camera_Site, sep = "")))%>%
  mutate(site = str_replace(site, "WW-W01", "WW-W1"),
         site = str_replace(site, "WW-W02", "WW-W2"),
         site = str_replace(site, "WW-W03", "WW-W3"),
         site = str_replace(site, "WW-W04", "WW-W4"),
         site = str_replace(site, "WW-W05", "WW-W5"),
         site = str_replace(site, "WW-W06", "WW-W6"),
         site = str_replace(site, "WW-W07", "WW-W7"),
         site = str_replace(site, "WW-W08", "WW-W8"),
         site = str_replace(site, "WW-W09", "WW-W9"))

#Who did not get named properly?
#From the readme we can ignore S9, S15, S19, S20, H1, H7
#Not sure about S26, S27, S28, S29, maybe just no data? 
cams%>%
  filter(!site %in% unique(dets$site))

unique(dets$site)

#Who do we not have coordinates for??
unique(dets$site) %in% unique(cams$site)


#But this is fine I think, now we can get the camera coordinates for all 115 sites
cams<-cams%>%
  rename(site = Camera_Site)%>%
  dplyr::select(site, Easting, Northing)
  
#Some stuff before the merge to check
n_distinct(dets$site) #should be 115
n_distinct(cams$site) #should be 125 (10 we arent using)

camop<-merge(dets, cams, by = "site")%>%
  select(site, Easting, Northing, start_date, end_date)


```

Preparing the independent detections
```{r}
#First we will make a new detection frame by merging the winter and summer dataframes
dets_all<-rbind(dets_winter, dets_summer)%>%
  select(site = Location, datetime, species = Species)

#Looks like there is a weird blank species? get rid of it
#I am going to leave all other species (including unknown stuff), as it will be up to the individual user to figure out how they want to deal with specifics
unique(dets_all$species)
dets_all<-dets_all%>%
  filter(species != "")

#Ok now, arrange all the data by site, species, and time
#And calculate a lag time for each subsequent species detection at each site
#THIS WILL LOOK SLIGHTLY STRANGE - this willmore data is missing the "seconds" from its datetime data, which is slightly annoying
#Should try to track down the original timelapse files
dets_all<-dets_all%>%
  arrange(site, species, datetime)%>%
  group_by(site, species)%>%
  mutate(timediff = as.numeric(difftime(datetime, lag(datetime),
                                        units = "mins")))

#This is our rolling window for "independence": we will count windows as independent if they are more than "mins" apart. Here we are using 30 minutes. 

mins<-30

#This loop assigns group ID based on independence. 
 dets_all$Event.ID <- 9999
 seq <- as.numeric(paste0(nrow(dets_all),0))
 seq <- round(seq,-(nchar(seq)))
 for (i in 2:nrow(dets_all)) {
   dets_all$Event.ID[i-1]  <- paste0("E",format(seq, scientific = F))
   if(is.na(dets_all$timediff[i]) | abs(dets_all$timediff[i]) > (mins)){
     seq <- seq + 1
   }
 }
 
 if(dets_all$timediff[nrow(dets_all)] < (mins)|
    is.na(dets_all$timediff[nrow(dets_all)])){
   dets_all$Event.ID[nrow(dets_all)] <- dets_all$Event.ID[nrow(dets_all)-1]
 } else{dets_all$Event.ID[nrow(dets_all)] <- paste0("E",format(seq+1, scientific = F))
 }
 
  #And then take only the first row of data for each independent event
dets_all<-dets_all%>%
   group_by(Event.ID)%>%
   slice(1)%>%
  arrange(site, species)


```


##############################################################################################################################
Preparing the covariate files

```{r}




```





And finally, write those outputs
```{r}
#output the detection file
write.csv(dets_all, file.path(output_directory, "willmore_30min_independent_detections.csv"), 
          row.names = FALSE) #Omit the row index (annoying)

#output the camera operability file
write.csv(camop, file.path(output_directory, "willmore_camop.csv"),
          row.names = FALSE)

```
